Supplementary Materials for \"COATi: statistical pairwise alignment of protein coding sequences\

### 2024-06-07 alignpair_letter.tex
Draft of final version after acceptence

\usepackage{url}
\usepackage{graphicx}

@@

% now start line numbers
\linenumbers

@@

Within protein-coding sequences, indels may occur in between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}). Gaps that occur after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively. Notably, phase-3 gaps occur between codons and are also known as phase-0 gaps \citep[e.g.][]{taylor2004occurrence}. Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2 which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}. While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps. Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce sub-optimal alignments and inflated estimates of sequence divergence
\citep[Fig.\ \ref{fig:aln};][]{redelings2007incorporating}.

% Figures are 86 mm or 178 mm wide
\begin{figure}[h!]
    \centering%
    \includegraphics[scale=1]{figures/fig-aln.pdf}
    \par
    \caption{
        Standard algorithms produce suboptimal alignments.
        (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
        (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
        %
        Indels in protein-coding sequences can be classified as having one of three difference phases and being one of two different types.
        Phases refer the the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
        Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the blue indel is type-I (an amino-acid indel only). The difference between an in-frame and a frameshift indel is not displayed.
        }
    \label{fig:aln}
\end{figure}

Bioinformatic pipelines need to be robust to variation in quality across genomic datasets because uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citep{estimates_schneider_2009, effect_fletcher_2010, hubisz2011error}.
While genomes for model organisms often get refined over many iterations and contain meticulously curated protein-coding sequences, 
genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citep{jackman2018tigmint}.
%
When comparative and functional genomics studies include data from non-model organisms, care must be taken to identify and manage such artifacts; however,
current alignment methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.

To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic data sets. We find that COATi generates biologically reasonable alignments and downstream inferences when applied to empirical data. We also find that COATi generates more accurate alignments when applied to semi-empirical benchmarks.

@@

In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{hein2001algorithm,holmes2001evolutionary,Lunter2005-hk,demaio2021cumulative}. An advantage of statistical alignment is that its parameters are derived from biological processes, allowing them to be estimated directly from data or extracted from previous studies. While approaches vary, a statistical aligner for a pair of sequences, $X$ and $Y$, typically finds an alignment, $Aln$, that maximizes the joint probability $P(Aln, X, Y)$ or samples alignments from the posterior $P(Aln | X, Y)$. This is typically performed using pairwise hidden Markov models \citep[pair-HMMs;][]{holmes2001evolutionary,redelings2007incorporating,problems_cartwright_2009}. Pair-HMMs are computational machines with two output tapes. Each tape represents one sequence, and a path through the pair-HMM represents an alignment of the two sequences. Conceptually, pair-HMMs generate two sequences from an unknown ancestor and can calculate the joint probability $P(Aln, X, Y)$ \citep{yoon_2009_hmm}.

@@

In order to use the COATi FST to align an output sequence against an input sequence, we first convert each sequence into an acceptor, represented as a linear transducer where the input and output symbols of each transition are identical and each transition represents one nucleotide of a sequence \citep{allauzen2007openfst}. By composing the input and output acceptors with the COATi FST, we generate a transducer of all possible alignments of the two sequences. Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment. If more that one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm. All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}. An example of an FST-based alignment can be found in Supplementary Materials, Figure S1.

@@

In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previous contain indels. We separated our 14,127 empirical sequence pairs into two sets: %
(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, ClustalΩ, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark set of known alignments. The phases of gaps were preserved as were the spacings of any clusters of gaps. Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if gap pattern did not fit the ungapped sequence pair.)

For example, consider the gapped alignment represented by the CIGAR string
%
``170M 3D 10M 6I 102M'' % Can use \allowbreak{} here instead of space
%
applied to an ungapped alignment of human and gorilla sequences that are 300 nucleotides long. First, we modified the CIGAR string to insert flexible lengths for any match segment that is 96 or more nucleotides long while preserving phase. This results in a new CIGAR string of
%
``98M *M 3D 10M 6I 96M *M'',
%
where * represents the locations that have flexible length. Considering only matches and deletions, this CIGAR string has 207 fixed nucleotides, leaving 93 nucleotides in the human sequence to be allocated to the flexible locations. Since there are two locations of flexible length, we drew one random break point uniformly while maintaining phase, producing the final CIGAR string of
%
``98M 12M 3D 10M 6I 96M 81M'',
%
which was used to add gaps into the target, ungapped sequence pair.
To apply deletions, we removed the corresponding nucleotides from the gorilla sequence, and to apply insertions, we added nucleotides to the gorilla sequence by sampling codons from their stationary frequency while respecting gap phases. The human sequence was left unchanged.
%
We generated a benchmark of 8261 sequence pairs with known alignments using this strategy.

@@

We aligned our benchmark alignments using our suite of aligners and quantified the similarity of each alignment that was produced to its respective benchmark. As above, we measured Kimura's 2-parameter distance \citep{kimura1980simple} and $K_A/K_S$ \citep{ka_ks_li_1993} for the benchmark and estimated alignments. Additionally, we quantified the error of estimated alignments using the $d_{seq}$ metric \citep{metrics_blackburne_whelan_2011}.

@@

Using $d_{seq}$ we quantified the number of perfect, imperfect, and best alignments each aligner produced. We define perfect alignments as alignments with a distance of zero to the benchmark alignment ($d_{seq} = 0$) or any alignment that is evolutionary equivalent to an alignment with a distance of zero. We included equivalency in our definition of perfect alignments to prevent the manner by which aligners break ties from influencing our results. Evolutionary equivalence was determined by scoring alignments using COATi's marginal model. Any alignment that had a score which matched the score of its benchmark alignment was considered perfect even if its distance to the benchmark was greater than zero.
%
Imperfect alignments are defined as alignments that are not perfect when another method successfully produces a perfect alignment for the same pair of sequences.
%
Best alignments are the alignments that the lowest distance $d_{seq}$ to the true alignment, including ties and equivalent alignments.
%
Taken together, these three statistics not only allow a direct comparison of aligners but also expose instances where all aligners fall short of achieving a perfect result.

@@

The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable for human-gorilla gene pairs, with means of 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a). Conversely, the means for ClustalΩ, MACSE, and MAFFT were much higher, 1.5\%--1.8\%. These larger means are due to the fact that the right tails of the distributions for ClustalΩ, MACSE, and MAFFT were larger than the other methods. As expected, the alignment methods that ``overaligned'' sequences (produced more mismatches and fewer gaps) also produced higher evolutionary distances on average (Tab.~\ref{tab:gaps-empirical} and Fig.~\ref{fig:k2p-empirical}a). The differences between the distances inferred by COATi and other methods also shows ClustalΩ, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
%
COATi's alignments produced significantly lower K2P distances than ClustalΩ, MACSE, MAFFT, and PRANK. Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$. Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test). Additional analysis of the empirical alignments can be found in Supplementary Methods, including results using p-distances, which did not differ from the results using K2P distances.

@@

The vast majority of sequence pairs (12,344) were identified as showing evidence of negative selection when they were aligned using COATi, ClustalΩ, MACSE, MAFFT, PRANK, and COATi-rev (Fig.~\ref{fig:sel-empirical}). Another 1087 sequence pairs were identified as showing evidence of positive selection when they were aligned using all 6 strategies. The most common pattern that showed disagreement between aligners were 91 sequence pairs that were identified as positively selected by COATi, PRANK, and COATi-rev. The most common singleton pattern was 64 sequence pairs identified as positively selected only by MAFFT. Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by by COATi and COATi-rev only.
%
In total, COATi identified 1,367 (\sim{}10\%) sequence pairs as showing evidence of positive selection. ClustalΩ identified 1,228. MACSE identified 1,183. MAFFT identified 1,340. And PRANK identified 1,349. Note that we did not explore whether any of these inferences were biologically or statistically significant.

@@

COATi produced better alignments than ClustalΩ, MACSE, MAFFT, and PRANK when evaluated by our benchmark data set. It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
%
COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}). All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests. (P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$). Notably, the average alignment error of the second best protocol was four times larger than COATi's. We also calculated the average distance of alignments between aligners and used principle coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}). COATi was the closest aligner to the benchmark, and other aligners tended to diverge from the benchmark in different directions. Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for each source of gap patterns (Supplementary Materials, Fig.~S6). While all aligners, except ClustalΩ, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.

@@

% Software comparison table
\begin{table}[!ht]
\definecolor{bestcolor}{RGB}{230,230,230}%
\centering
\caption{COATi generates better alignments than other alignment algorithms on semi-empirical benchmarks.\textsuperscript{1}\label{table:comp}}
%
% Copied from tables/benchmarks.tex
% footnotes and \addlinespace are manually added
\begin{tabular}{rccccc}
  \toprule
 & COATi & ClustalΩ & MACSE & MAFFT & PRANK \\ 
  \midrule
Average Alignment Error ($d_{seq}$) & \cellcolor{bestcolor}0.23\% & 2.65\% & 1.37\% & 1.67\% & 1.02\% \\
\addlinespace 
  Number of Best\textsuperscript{2} Alignments & \cellcolor{bestcolor}7050 & 3988 & 5226 & 5918 & 6128 \\ 
  Number of Perfect\textsuperscript{2} Alignments & \cellcolor{bestcolor}6795 & 3954 & 5138 & 5840 & 5860 \\ 
  Number of Imperfect\textsuperscript{2} Alignments & \cellcolor{bestcolor} 850 & 3691 & 2507 & 1805 & 1785 \\
  \addlinespace 
  RMSE for K2P Distances & \cellcolor{bestcolor}0.000694 & 0.0920 & 0.0528 & 0.0561 & 0.0305 \\ 
  Overestimated K2P Distances & \cellcolor{bestcolor}10.3\% & 52.7\% & 38.0\% & 26.2\% & 25.6\% \\ 
  \addlinespace
  F\textsubscript{1} Score for Negative Selection & \cellcolor{bestcolor}99.8\% & 97.6\% & 98.4\% & 98.3\% & 99.1\% \\ 
  F\textsubscript{1} Score for Positive Selection & \cellcolor{bestcolor}97.8\% & 77.2\% & 84.9\% & 85.5\% & 91.8\% \\ 
   \bottomrule
   \addlinespace
\multicolumn{6}{p{40em}}{\textsuperscript{1}Total number of sequence pairs in the benchmark was 8,261.}\\
\multicolumn{6}{p{40em}}{\textsuperscript{2}Best alignments have the lowest $d_{seq}$ (including equivalent alignments). Perfect alignments have $d_{seq} = 0$ (including equivalent alignments). Imperfect alignments are alignments that are not perfect when at least one method found a perfect alignment. }\\
\end{tabular}
\end{table}

\begin{figure}[h!]
    \centering%
    \includegraphics{figures/fig-dseq-benchmarks.pdf}
    \par
    \caption{COATi's alignments were closer to the semi-empirical benchmark than other methods. A principal coordinates analysis (PCoA) of the average alignment distance ($d_{seq}$) between alignments generated by different methods.
%
    \label{fig:dseq-benchmarks}}
\end{figure}

@@

We have developed a statistical alignment program using finite-state transducers called COATi. On empirical data, COATi produced alignments that provided reasonable biological inferences. On semi-empirical benchmarks COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model. Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps. Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences between humans and gorillas are phase-3 indels. Therefore alignment strategies that only support phase-3 gaps in protein-coding sequences are sub-optimally aligning the 59\% of indels that are phase-1 and phase-2.

In addition to supporting all three phases of gaps, COATi's advantage over the other aligners that we tested is at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings. Different data sets may show different results. However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}. This contrasts with other aligners that have limited ability to customize parameters and/or use parameters without biological interpretations.

@@

We can look at distances between alignments generated by different aligners to understand the diversity of approaches use by our alignment protocols. Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7). Different aligners tended to spread out from one another and in different directions from the benchmark. This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''. While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK did well on ClustalΩ patterns. COATi did well on all patterns.

This study has focused on COATi's module for estimating pairwise alignments; however, COATi has additional tools, including modules to sample pairwise alignments from alignment space and generate multiple-sequence alignments. We have used the sampling module in a large study \citep{zhu2022profiling} and believe that it is ready to be used in further studies. On the other hand, COATi's multiple-sequence alignment algorithm is still in early development and needs additional work to ensure that it produces high quality alignments. Currently, COATi generates multiple sequence alignments by aligning non-reference sequences against a reference sequence and collapsing insertions following a guide tree.

@@

COATi is under active development, and future work will enhance the multiple-sequence aligner to refine the initial alignment according to a scoring function consistent with COATi's FST model. We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighing gap openings to reflect known selection on indel phases \citep{zhu2022profiling}. We plan on improving its alignment sampling capabilities, as well as implement new models for aligning long-read sequences of genes against reference genomes. Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.

>>>>

\usepackage{url}
\urlstyle{same}
\usepackage{graphicx}

@@

% now start line numbers
% \linenumbers

@@

Within protein-coding sequences, indels may occur between any pair of adjacent nucleotides, and therefore, gaps in alignments of natural sequences may occur both between and within codons (Fig.\ \ref{fig:aln}). In our nomenclature, gaps that start after the first, second, and third positions of a codon are known as phase-1, phase-2, and phase-3 gaps, respectively. Here, gaps that occur between codons are phase-3 gaps whereas they may be known as phase-0 gaps in earlier studies \cite[e.g.][]{taylor2004occurrence}. Indels that occur within codons can produce an amino-acid change in addition to adding or removing nucleotides (Fig.\ \ref{fig:aln}), and due to the structure of the genetic code, phase-1 gaps are less common than phase-2, which are less common than phase-3 \citep{taylor2004occurrence, zhu2022profiling}. While all three phases of gaps occur in natural sequences, alignments performed in amino-acid or codon space force all gaps to be phase-3 gaps. Because only about 42\% of indels are phase-3 \citep{taylor2004occurrence, zhu2022profiling}, this mismatch between aligner assumptions and biology can produce suboptimal alignments and inflated estimates of sequence divergence
\citep[Fig.\ \ref{fig:aln};][]{redelings2007incorporating}.

% Figures are 86 mm or 178 mm wide
\begin{figure}[h!]
    \centering%
    \includegraphics[scale=1]{figures/fig-aln.pdf}
    \par
    \caption{
        Standard algorithms produce suboptimal alignments.
        (a) shows the true alignment of an ancestor sequence (A) and a descendant sequence (D).
        (b--d) are the results of different aligners. Nucleotide mismatches are highlighted in red.---Notably, COATi is the only aligner able to retrieve the biological alignment in this example.---%
        %
        Indels in protein-coding sequences can be classified as having one of three different phases and being one of two different types.
        Phases refer to the location of the gap with respect to the reading frame, while types refer to the consequence of the indel.
        Phase-1, phase-2, and phase-3 indels are shown in blue, orange, and green respectively.
        Additionally, the orange indel is type-II (an amino-acid indel plus an amino-acid change) while the blue indel is type-I (an amino-acid indel only). The difference between an in-frame and a frameshift indel is not displayed.
        }
    \label{fig:aln}
\end{figure}

Bioinformatic pipelines need to be robust to variation in quality across genomic datasets because uncorrected errors in the alignment stage can lead to erroneous results in comparative and functional genomic studies \citep{estimates_schneider_2009, effect_fletcher_2010, hubisz2011error}.
While genomes for model organisms are often refined over many iterations and contain meticulously curated protein-coding sequences, 
genomes for non-model organisms might only receive partial curation and typically have lower quality sequences and annotations.
These genomes often lack the amount of sequencing data needed to fix artifacts, including missing exons, erroneous mutations, and indels \citep{jackman2018tigmint}.
%
When comparative and functional genomics studies include data from non-model organisms, care must be taken to identify and manage such artifacts; however,
current alignment methods are ill-equipped to handle common artifacts in genomic data, requiring costly curation practices that discard significant amounts of information.

To address current limitations of alignment software to accurately align protein-coding sequences, we present COATi, short for COdon-aware Alignment Transducer, a pairwise statistical aligner that incorporates evolutionary models for protein-coding sequences and is robust to artifacts present in modern genomic datasets. We find that COATi generates biologically reasonable alignments and downstream inferences when applied to empirical data. We also find that COATi generates more accurate alignments when applied to a semi-empirical benchmark dataset.

@@

In statistical alignment, sequence alignments are scored based on a stochastic model, typically derived from molecular evolutionary processes \citep{hein2001algorithm,holmes2001evolutionary,Lunter2005-hk,Holmes2020,demaio2021cumulative}. An advantage of statistical alignment is that its parameters are derived from biological processes, allowing them to be estimated directly from data or extracted from previous studies. While approaches vary, a statistical aligner for a pair of sequences, $X$ and $Y$, typically finds an alignment, $Aln$, that maximizes the joint probability $P(Aln, X, Y)$ or samples alignments from the posterior $P(Aln | X, Y)$. This is typically performed using pairwise hidden Markov models \citep[pair-HMMs;][]{holmes2001evolutionary,redelings2007incorporating,problems_cartwright_2009}. Pair-HMMs are computational machines with two output tapes. Each tape represents one sequence, and a path through the pair-HMM represents an alignment of the two sequences. Conceptually, pair-HMMs generate two sequences from an unknown ancestor and can calculate the joint probability $P(Aln, X, Y)$ \citep{yoon_2009_hmm}.

@@

In order to use the COATi FST to align an output sequence against an input sequence, we first convert each sequence into an acceptor, represented as a linear transducer where the input and output symbols of each transition are identical and each transition represents one nucleotide of a sequence \citep{allauzen2007openfst}. By composing the input and output acceptors with the COATi FST, we generate a transducer of all possible alignments of the two sequences. Any path through this FST represents a pairwise alignment, while the shortest path (by weight) corresponds to the best alignment. If more than one optimal alignment exists, ties are broken according to the implementation of the shortest-path algorithm. All FST operations in COATi, including model development, composition, search for the shortest path, and other optimization algorithms, are performed using the C++ openFST library \citep{allauzen2007openfst}. An example of an FST-based alignment can be found in Supplementary Materials, Figure S1.

@@

In order to compare COATi against other aligners on realistic datasets with known alignments, we developed a procedure to introduce realistic gap patterns into human-gorilla orthologous gene pairs that did not previously contain indels. We separated our 14,127 empirical sequence pairs into two sets: %
(1) 8261 sequence pairs that did not contain any gaps when aligned by COATi FST, ClustalΩ, MACSE, MAFFT, and PRANK, and (2) 5821 sequence pairs for which at least one of these five aligners added gaps.
(45 pairs that could not be aligned by at least one aligner were dropped.) We identified gap patterns from the sequence pairs that were aligned with gaps and randomly introduced them into the ungapped sequence pairs. We used an equal number of randomly sampled gap patterns from each aligner to produce a benchmark dataset of known alignments. The phases of gaps were preserved as were the spacings of any clusters of gaps. Segments of matches that were 96 nucleotides or longer were allowed to change length to accommodate the length of the ungapped sequence pair. (This criteria was recursively lowered if the gap pattern did not fit the ungapped sequence pair.)

For example, consider the gapped alignment represented by the CIGAR string
%
``170M 3D 10M 6I 102M'' % Can use \allowbreak{} here instead of space
%
applied to an ungapped alignment of human and gorilla sequences that are 300 nucleotides long. First, we modified the CIGAR string to insert flexible lengths for any match segment that is 96 or more nucleotides long while preserving phase. This results in a new CIGAR string of
%
``98M *M 3D 10M 6I 96M *M'',
%
where * represents the locations that have flexible length. Considering only matches and deletions, this CIGAR string has 207 fixed nucleotides, leaving 93 nucleotides in the human sequence to be allocated to the flexible locations. Since there are two locations of flexible length, we drew one random break point uniformly while maintaining phase, producing the final CIGAR string of
%
``98M 12M 3D 10M 6I 96M 81M'',
%
which was used to add gaps into the target, ungapped sequence pair.
To apply deletions, we removed the corresponding nucleotides from the gorilla sequence, and to apply insertions, we added nucleotides to the gorilla sequence by sampling codons from their stationary frequency while respecting gap phases. The human sequence was left unchanged.
%
We generated a benchmark dataset of 8261 sequence pairs with known alignments using this strategy.

@@

We aligned our benchmark alignments using our suite of aligners and quantified the similarity of each estimated alignment to its respective benchmark. As above, we measured Kimura's 2-parameter distance \citep{kimura1980simple} and $K_A/K_S$ \citep{ka_ks_li_1993} for the benchmark alignments and estimated alignments. Additionally, we quantified the error of estimated alignments using the $d_{seq}$ metric \citep{metrics_blackburne_whelan_2011}.

@@

Using $d_{seq}$ we quantified the number of perfect, imperfect, and best alignments each aligner produced. We define perfect alignments as alignments with a distance of zero to the benchmark alignment ($d_{seq} = 0$) or any alignment that is evolutionary equivalent to an alignment with a distance of zero. We included equivalency in our definition of perfect alignments to prevent the manner by which aligners break ties from influencing our results. Evolutionary equivalence was determined by scoring alignments using COATi's marginal model. Any alignment that had a score which matched the score of its benchmark alignment was considered perfect even if its distance to the benchmark alignment was greater than zero.
%
Imperfect alignments are defined as alignments that are not perfect when another method successfully produces a perfect alignment for the same pair of sequences.
%
Best alignments are the alignments that the lowest distance $d_{seq}$ to the true alignment, including ties and equivalent alignments.
%
Taken together, these three statistics not only allow a direct comparison of aligners but also expose instances where all aligners fall short of achieving a perfect result.

@@

The distribution of evolutionary distances for COATi, COATi-rev, and PRANK were biologically reasonable for human-gorilla gene pairs, with means of 0.8\%--0.9\% (Fig.~\ref{fig:k2p-empirical}a). Conversely, the means for ClustalΩ, MACSE, and MAFFT were much higher, 1.5\%--1.8\%. These larger means are due to the fact that the right tails of the distributions for ClustalΩ, MACSE, and MAFFT were larger than the other methods. As expected, the alignment methods that ``overaligned'' sequences (produced more mismatches and fewer gaps) also produced higher evolutionary distances on average (Tab.~\ref{tab:gaps-empirical} and Fig.~\ref{fig:k2p-empirical}a). The differences between the distances inferred by COATi and other methods also shows ClustalΩ, MACSE, and MAFFT producing higher distance estimates for a subset of sequences (Fig.~\ref{fig:k2p-empirical}b).
%
COATi's alignments produced significantly lower K2P distances than ClustalΩ, MACSE, MAFFT, and PRANK. Single-tailed Wilcoxon signed-rank tests on matched data produced p-values $< 2.2\times10^{-16}$. Additionally, COATi's K2P distances did not depend on whether human or gorilla was used as the reference sequence (p-value of $0.4212$ for a two-sided test). Additional analyses of the empirical alignments can be found in Supplementary Methods, including results using p-distances, which did not differ from the results using K2P distances.

@@

The vast majority of sequence pairs (12,344) were identified as showing evidence of negative selection when they were aligned using COATi, ClustalΩ, MACSE, MAFFT, PRANK, and COATi-rev (Fig.~\ref{fig:sel-empirical}). Another 1087 sequence pairs were identified as showing evidence of positive selection when they were aligned using all 6 strategies. The most common pattern that showed disagreement between aligners were 91 sequence pairs that were identified as positively selected by COATi, PRANK, and COATi-rev. The most common singleton pattern was 64 sequence pairs identified as positively selected only by MAFFT. Notably, 36 sequence pairs were identified as positively selected and 9 sequence pairs were identified as negatively selected by COATi and COATi-rev only.
%
In total, COATi identified 1,367 (\sim{}10\%) sequence pairs as showing evidence of positive selection. ClustalΩ identified 1,228. MACSE identified 1,183. MAFFT identified 1,340. And PRANK identified 1,349. Note that we did not explore whether any of these inferences were biologically or statistically significant.

@@

COATi produced better alignments than ClustalΩ, MACSE, MAFFT, and PRANK when evaluated by our benchmark dataset. It produced more best alignments, more perfect alignments, and less imperfect alignments (Tab.~\ref{table:comp}).
%
COATi was significantly more accurate (lower $d_{seq}$) at inferring the benchmark alignments compared to the other methods (Tab.~\ref{table:comp}). All p-values were significant according to the one-tailed, paired Wilcoxon signed-rank tests. (P-value of COATi vs PRANK was $5.1 \times 10^{-8}$, and p-values of COATi vs the other aligners were $< 2.2 \times 10^{-16}$). Notably, the average alignment error of the second best protocol was four times larger than COATi's. We also calculated the average distance of alignments between aligners and used principal coordinate analysis (PCoA) to project the resulting distance matrix into two dimensions (Fig.~\ref{fig:dseq-benchmarks}). COATi was the closest aligner to the benchmark dataset, and other aligners tended to diverge from the benchmark dataset in different directions. Because benchmark alignments were derived from gap patterns estimated by different aligners, we also performed separate PCoA analyses for every source of gap patterns (Supplementary Materials, Fig.~S6). While all aligners, except ClustalΩ, do exceptionally well on their own patterns, COATi does exceptionally well on every set of patterns.

@@

% Software comparison table
\begin{table}[!ht]
\definecolor{bestcolor}{RGB}{230,230,230}%
\centering
\caption{COATi generates better alignments than other alignment algorithms on a semi-empirical benchmark dataset.\textsuperscript{1}\label{table:comp}}
%
% Copied from tables/benchmarks.tex
% footnotes and \addlinespace are manually added
\begin{tabular}{rccccc}
  \toprule
 & COATi & ClustalΩ & MACSE & MAFFT & PRANK \\ 
  \midrule
Average Alignment Error ($d_{seq}$) & \cellcolor{bestcolor}0.23\% & 2.65\% & 1.37\% & 1.67\% & 1.02\% \\
\addlinespace 
  Number of Best\textsuperscript{2} Alignments & \cellcolor{bestcolor}7050 & 3988 & 5226 & 5918 & 6128 \\ 
  Number of Perfect\textsuperscript{2} Alignments & \cellcolor{bestcolor}6795 & 3954 & 5138 & 5840 & 5860 \\ 
  Number of Imperfect\textsuperscript{2} Alignments & \cellcolor{bestcolor} 850 & 3691 & 2507 & 1805 & 1785 \\
  \addlinespace 
  RMSE for K2P Distances & \cellcolor{bestcolor}0.000694 & 0.0920 & 0.0528 & 0.0561 & 0.0305 \\ 
  Overestimated K2P Distances & \cellcolor{bestcolor}10.3\% & 52.7\% & 38.0\% & 26.2\% & 25.6\% \\ 
  \addlinespace
  F\textsubscript{1} Score for Negative Selection & \cellcolor{bestcolor}99.8\% & 97.6\% & 98.4\% & 98.3\% & 99.1\% \\ 
  F\textsubscript{1} Score for Positive Selection & \cellcolor{bestcolor}97.8\% & 77.2\% & 84.9\% & 85.5\% & 91.8\% \\ 
   \bottomrule
   \addlinespace
\multicolumn{6}{p{40em}}{\textsuperscript{1}Total number of sequence pairs in the benchmark dataset was 8,261.}\\
\multicolumn{6}{p{40em}}{\textsuperscript{2}Best alignments have the lowest $d_{seq}$ (including equivalent alignments). Perfect alignments have $d_{seq} = 0$ (including equivalent alignments). Imperfect alignments are alignments that are not perfect when at least one method found a perfect alignment. }\\
\end{tabular}
\end{table}

\begin{figure}[h!]
    \centering%
    \includegraphics{figures/fig-dseq-benchmarks.pdf}
    \par
    \caption{COATi's alignments were closer to the semi-empirical benchmark dataset than other methods. A principal coordinates analysis (PCoA) of the average alignment distance ($d_{seq}$) between alignments generated by different methods.
%
    \label{fig:dseq-benchmarks}}
\end{figure}

@@

We have developed a statistical alignment program using finite-state transducers called COATi. On empirical data, COATi produced alignments that provided reasonable biological inferences. On a semi-empirical benchmark dataset, COATi was more accurate than the other aligners/alignment strategies that we tested: ClustalΩ's amino-acid model, MACSE's amino-acid aware nucleotide model, MAFFT's DNA model, and PRANK's codon model. Unlike standard codon or amino-acid alignment strategies, COATi supports aligning protein-coding sequences using all three phases of gaps. Consistent with \cite{taylor2004occurrence} and \cite{zhu2022profiling}, our COATi results indicate that only 41\% indels in protein-coding sequences are phase-3 indels between humans and gorillas. Therefore, alignment strategies that only support phase-3 gaps in protein-coding sequences are suboptimally aligning the 59\% of indels that are phase-1 and phase-2.

In addition to supporting all three phases of gaps, COATi's advantage over the other aligners was at least partially due to its default parameters fitting the human-gorilla coding sequences better than the other aligners' default settings. Different data sets may show different results. However, because COATi uses a statistical model with biologically meaningful parameters, users can easily optimize COATi by using parameters estimated by previous studies or estimating parameters themselves \citep[e.g.][]{zhu2022profiling}. This contrasts with other aligners that have limited ability to customize parameters and/or use parameters without biological interpretations.

@@

We can look at distances between alignments generated by different aligners to understand the diversity of approaches used by our alignment protocols. Aside from the cluster of different COATi methods, we detected no other clusters in our analysis (Supplemental Materials Fig.~S7). Different aligners tended to spread out from one another and in different directions from the benchmark dataset. This indicates that each protocol that we utilized approached alignment inference from a different section of ``aligner space''. While each aligner, except ClustalΩ, performed well on its own gap patterns, MACSE also did well on ClustalΩ and MAFFT gap patterns, and PRANK did well on ClustalΩ patterns. COATi did well on all patterns.

This study has focused on COATi's module for estimating pairwise alignments; however, COATi has additional tools, including a module to sample pairwise alignments from alignment space and a module to generate multiple-sequence alignments. We have used the sampling module in a large study \citep{zhu2022profiling} and believe that it is ready to be used in further studies. On the other hand, COATi's multiple-sequence alignment algorithm is still in early development and needs additional work to ensure that it produces high quality alignments. Currently, COATi generates multiple sequence alignments by aligning non-reference sequences against a reference sequence and collapsing insertions following a user-provided guide tree.

@@

COATi is under active development, and future work will enhance the multiple-sequence aligner to refine the initial alignment according to a scoring function consistent with COATi's FST model. We also plan on extending COATi to support more complex gap models, e.g.\ mixtures of single-nucleotide and triple-nucleotide indel models and weighting gap openings to reflect known selection on indel phases \citep{zhu2022profiling}. We also plan on improving its alignment sampling capabilities, as well as implementing new models for aligning long-read sequences of genes against reference genomes. Our goal is to develop COATi into a user-friendly suite of tools that will allow researchers to analyze more data with higher accuracy and facilitate the study of important biological processes that shape genomic data.
